Non-word repetition in Yélî Dnye

### 2020-10-05 manuscript.Rmd
minor

cohensd=function(m,sd){round(m/sd,3)}
cohensd2=function(m1,sd1,m2,sd2){round((m1-m2)/sqrt((sd1^2+sd2^2)/2),3)}
```

@@

- double check my frequency entries in segments.xlsx using the search function in https://phoible.org/parameters; take a look at http://phoible.github.io/conventions/ in case you see something about double articulation (I didn't find tp or lBj)
<!--M2A: I think you mentioned you might have someone do this check, but let me know if I misunderstood!-->
- draft discussion

<!-- M2A: regarding demo, it looks like you have more mateds than birth orders because family data was systematically not collected for Pu. kids and was missing for three kids tested in No. who were actually from other hamlets. New xslx files sent via email! I left age recomputing to you since I was confused about which file would be used.-->
<!-- M2A: regarding NAs in stim_seg_freq, these appear for 'lv' stimuli because, in the 'segments' table, freq_corpus and freq_corpus_types are NA for that grapheme). The 'segments' table has NAs for 'lv' because it is not included in segment-counts.txt or segment-counts-types.txt. Note that that these NAs are excluded when adding mean item frequency to the 'stims' table (wrangling.R/l.136 & l.145), but this has a pretty big impact since those are all two-segment words. Not sure how avoidable this is, going back to our discussion before about not wanting to ad arbitrarily small frequencies. Thoughts? -->

@@

NWR has been used to seek answers to a variety of theoretical questions, including what  the links between phonology, working memory, and the lexicon are [@bowey2001nonword], and to assess the extent to which apparent phonological constraints found in the lexicon affect online production [@gallagher2014identity]. It is also frequently used for applications, notably  as a diagnostic for language delays and disorders [@estes2007differences]. Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments,  particularly in Europe [@meir2016disentangling;@is08042009language]. However, it has been seldom used outside of Europe and North America [with exceptions including @gallagher2014identity and @cristia2020infant].

<!--M2A: I feel like there's a bit of a jump here from explaining what NWR is and what it's used for to saying we used it in the current study to test three things. We should probably have a paragraph justifying its use in this context, no?-->
<!-- a2m I added a sentence, is this what you were thinking of? -->
<!-- M2A: Makes sense. I was thinking we could already hint at YD as interesting in particular bc of the phonology and not simply because it's a random one of many previously untested populations. I don't mind adding something like this in once I'm done with the demo check.-->

@@

Most speakers of the language grow up speaking it monolingually until they begin attending school around the age of 7; 
<!-- a2m no "at" here? *at* around the age? -->
<!-- M2A: my ear strongly prefers it without the "at", but it could go both ways. -->
school instruction is in English. While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---, where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing]. Children in these multilingual families grow up speaking Yélî Dnye and English, Tok Pisin, and/or some other language(s) from the region.

@@

While work on Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2020early; @liszkowski2012prelinguistic], our knowledge of children's linguistic development is quite limited, and research into their phonological development in particular has only just begun [e.g., Bram's thesis; MC's CFH data]. What we do know about the early language environment is that, while child-directed speech is relatively infrequent throughout the day [@casillas2020early], shared caregiving practices and a near-universal fondness for social interaction with young children ensure that speech directed to children comes from all types of speakers: women, men, and other children [@bunceURcrosscultural; @casillas2020early]. While speech from women predominates in young children's speech environments on Rossel, as it does elsewhere [@bergelsoncasillas2019what; @bunceURcrosscultural], there is a significant and marked increase in child-directed speech from other children as infants get older [@bunceURcrosscultural; @casillas2020early]. This increase in child-directed speech from other children is attributed to the fact that, starting around age 2, children often spend large swaths of the day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 people) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].
<!--M2A: I think a nice way to end this bit would be very briefly (~1-sentence) linking this input scenario to the relevance of looking at NWR (à la Cristia 2020 DevRev). What do you think?-->
<!-- a2m maybe we want to bring that up in the discussion? I haven't found a good place to introduce that paper, in part because right now we have these three purposes that seem sort of unconnected with the goals of that paper... -->
<!-- M2A: fine with me!-->

@@

The basic structure of NWR involves the participant hearing a non-word, like "bilik", which they need to repeat back without changing anything that is contrastive in the language. For instance, in English, a response of "bilig" or "pilik" will be scored as incorrect; a response "biilik", where the vowel is lengthened without change of quality would be scored as correct, because English does not have a pure length contrast. 
<!-- a2m added explanation of correct and incorrect for nwr task -->
<!--M2A: thanks!-->

@@

- Does the frequency of sounds across languages predict NWR? Are rarer sounds more often substituted by commoner sounds?
- How does score change as a function of item length in number of syllables?
- Is individual variation attributable to child age, sex, birth order, monolingual status, and/or parental education?
<!-- a2m we don't actually look at this: Are rarer sounds more often substituted by commoner sounds? -- shall we remove it? -->
<!--M2A: I'm imagining a reviewer who's knowlegeable about phonology and child development but who hasn't encountered NWR before. I think it would be beneficial to provide a concrete example of a nonword and an inexact repetition above so that they can conceive of what these questions imply (analytically) before we get into the methodological weeds.-->
<!-- a2m good point, but I think it goes in the previous section so I added it there -->
<!--M2A: thanks!-->

@@

# a2m I get a warning:
#   Warning message:
# Vectorized input to `element_text()` is not officially supported.
# Results may be unexpected or may change in future versions of ggplot2. 
# M2A: removed the vector since it wasn't needed; no more complaint!

@@

We visited four different hamlets once, and attempted to test all eligible children present at the time, to prevent the items 'spreading' through hearsay. In the first hamlet, we tested children in five different places, with some children being tested inside a house and others tested on the veranda. The complete list of places and the ways in which they met the desiderata mentioned above can be found in the raw data, available from online supplementary materials.
 <!--M2A: I didn't look for the SM yet, but probably best not to provide exact hamlet names-->
<!--a2m: good point! fixed-->
<!-- M2A: thanks! -->

@@

The first phase was making sure the child understood the task. This was explained orally and the first training item was presented. Often, children froze and did not say anything. If this happened, then we followed this procedure. First, the informant insisted. If the child still did not say anything, the informant asked the child to repeat a real word, and another, and another. If the child could repeat these correctly, then we provided the recorded training item over headphones again. Most children successfully started repeating the items presented over headphones at this point; a few further needed the local informant to model the behavior (i.e., they would hear the item again, and she would say it; then we would play it again, and ask the child to say it). A small minority still failed to repeat the item after hearing it over headphones. If that occurred, we tried with the second training item, at which point some children got it and could continue. A small minority, however, failed to repeat this one, as well as the third training item, in which case we stopped the test altogether (see Participants section for exclusions). 
<!--M2A: do you know how many these were? oh! it's below. then maybe we can just put a (see Participants below), etc.-->
<!-- a2m done -->
<!-- M2A: thanks! -->

@@

read.csv("NWR-demo.csv",header=T)->demo
#colnames(demo)

#if family is missing, attribute each child to a different one
demo$familyID=as.character(demo$familyID)
demo$familyID[is.na(demo$familyID)]<-paste("k",demo$ID[is.na(demo$familyID)])

#fix sex
demo$Sex[demo$Sex=="F"]<-"G"

@@

A total `r dim(demo)[1]` children were tested, from `r length(levels(factor(demo$familyID)))` of families, in five hamlets. Some children could not be included for the following reasons: refused participation or failed to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn't repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`). In addition, `r sum(demo$Age >12 | demo$included=="too old",na.rm=T)` teenagers were tested to put younger children at ease; their data is not included in analyses below. The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged `r round(mean(inc$age.rounded),2)` years (range `r round(min(inc$age.rounded),2)`-`r round(max(inc$age.rounded),2)` years). Among these, there were `r sum(inc$monolingual=="yes", na.rm=T)` children exposed only to Yélî in the home, `r sum(inc$monolingual=="no", na.rm=T)` children who were also exposed to another language in the home, and `r sum(is.na(inc$monolingual))` for whom this information is missing. Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years; `r sum(is.na(inc$matEd))` children had this information missing).[^1] In terms of birth order, `r birthorder.tab[1]` were first borns, `r birthorder.tab[2]` second,  `r birthorder.tab[3]` third,  `r birthorder.tab[4]` forth,  `r birthorder.tab[5]` fifth, `r birthorder.tab[6]` sixth, and  `r birthorder.tab[7]` did not have this information.

@@

```{r overall}
# M2A: This excludes a bunch of kids (anyone with NA on any of these participant features),
# whittling the summary down to 24 from 42 -- is that intended?
# Also all phon_score is 100 and all NLD is 0; does that stike you as fishy?
#a2m fixed and fixed!
# M2A: thanks!
aggregate(trnsc$correct*100,by=list(trnsc$id),mean)->overall
aggregate(trnsc$phon_score*100,by=list(trnsc$id),mean)->phon_score
aggregate(trnsc$nld*100,by=list(trnsc$id),mean)->nld
overall=cbind(overall,phon_score$x,nld$x)
overall=merge(overall, demo[,c("ID","Sex","Age","birthOrder","matEd")], by.x="Group.1",by.y="ID", all.x=T)
colnames(overall)<-c("ID","Word","Phoneme","NLD","Sex","Age","BrthO","matEd")

@@

<!-- a2m should we point out that the coefficient for cross-ling frequency is reduced, so indirectly corpus frequency does explain a little bit of variance? Or making things too complicated, and just say "so corpus frequency does not explain away these results" -->
<!-- M2A: let's not go into it. because of the way glmer treats collinearity (https://psycnet.apa.org/record/2014-08495-005), it gets complicated quickly in even giving an accurate report of what's going on -->

### Qualitative analyses of NWR mispronunciations

<!-- a2m I realized this section did not belong together with the xling freq section above - are you ok with this title? -->
<!-- M2A: The title is a little long and maybe less informative than it could be. How about something like, "Patterns in NWR mispronunciations"-->

@@

<!-- a2m I added info above - is it weird to talk about this here? -->
<!-- M2A: I don't think it's a weird place to add the idea, though I might shorten it a bit when I go through on a writing pass :) -->

@@

```{r bylength}
# M2A: this reduces the total number of kids represented from 42 to 40
#a2m fixed
# M2A: thanks!
aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$type),mean)->by_length
aggregate(trnsc$phon_score*100,by=list(trnsc$id,trnsc$type),mean)->phon_score
aggregate(trnsc$nld*100,by=list(trnsc$id,trnsc$type),mean)->nld
by_length=cbind(by_length,phon_score$x,nld$x)
by_length=merge(by_length, demo[,c("ID","Sex","Age","birthOrder","matEd")], by.x="Group.1",by.y="ID", all.x=T)
colnames(by_length)<-c("ID","Length","Word","Phoneme","NLD","Sex","Age","BrthO","matEd")

aggregate(trnsc$correct*100,by=list(trnsc$type),mean)->by_length_no_child
colnames(by_length_no_child)<-c("Length","Word")
by_length_no_child=by_length_no_child[-1,]
length_r=cor.test(by_length_no_child$Length,by_length_no_child$Word)
```

@@

**(MC: but we can try and do a cursory analysis based on the corpora we have from Steve and my transcription of naturalistic interactions!)**

@@

>>>>

cohensd=function(m,sd){round(m/sd,3)}
cohensd2=function(m1,sd1,m2,sd2){round((m1-m2)/sqrt((sd1^2+sd2^2)/2),3)}

@@

- double check my frequency entries in segments.xlsx using the search function in https://phoible.org/parameters; take a look at http://phoible.github.io/conventions/ in case you see something about double articulation (I didn't find tp or lBj)
<!--M2A: I think you mentioned you might have someone do this check, but let me know if I misunderstood!-->
<!-- a2m we then decided we'd both feel more certain if you did... -->

<!-- M2A: regarding demo, it looks like you have more mateds than birth orders because family data was systematically not collected for Pu. kids and was missing for three kids tested in No. who were actually from other hamlets. New xslx files sent via email! I left age recomputing to you since I was confused about which file would be used.-->
<!-- a2m thanks!!! now inc'ed -->

<!-- M2A: regarding NAs in stim_seg_freq, these appear for 'lv' stimuli because, in the 'segments' table, freq_corpus and freq_corpus_types are NA for that grapheme). The 'segments' table has NAs for 'lv' because it is not included in segment-counts.txt or segment-counts-types.txt. Note that that these NAs are excluded when adding mean item frequency to the 'stims' table (wrangling.R/l.136 & l.145), but this has a pretty big impact since those are all two-segment words. Not sure how avoidable this is, going back to our discussion before about not wanting to ad arbitrarily small frequencies. Thoughts? -->
<!-- a2m So your fear is that frequency is overestimated in these cases, right? Perhaps we can assign the missing segments the lowest observed frequencies? -->

- draft discussion

@@

NWR has been used to seek answers to a variety of theoretical questions, including what  the links between phonology, working memory, and the lexicon are [@bowey2001nonword], and to assess the extent to which apparent phonological constraints found in the lexicon affect online production [@gallagher2014identity]. It is also frequently used for applications, notably  as a diagnostic for language delays and disorders [@estes2007differences]. Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments,  particularly in Europe [@meir2016disentangling;@is08042009language]. However, it has been seldom used outside of Europe and North America [with exceptions including @gallagher2014identity and @cristia2020infant], and mostly with languages having phonological inventories that are small to moderate in size [@maddieson2005correlating].

<!--M2A: I feel like there's a bit of a jump here from explaining what NWR is and what it's used for to saying we used it in the current study to test three things. We should probably have a paragraph justifying its use in this context, no?-->
<!-- a2m I added a sentence, is this what you were thinking of? -->
<!-- M2A: Makes sense. I was thinking we could already hint at YD as interesting in particular bc of the phonology and not simply because it's a random one of many previously untested populations. I don't mind adding something like this in once I'm done with the demo check.-->
<!-- a2m added ", and mostly with languages having phonological inventories that are small to moderate in size" + notice the " with an unusually dense phonological inventory"; my feeling is that it may interrupt the flow to add a lot more here... don't you think? -->

@@

Most speakers of the language grow up speaking it monolingually until they begin attending school around the age of 7; 
school instruction is in English. While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---, where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing]. Children in these multilingual families grow up speaking Yélî Dnye and English, Tok Pisin, and/or some other language(s) from the region.

@@

While work on Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2020early; @liszkowski2012prelinguistic], our knowledge of children's linguistic development is quite limited, and research into their phonological development in particular has only just begun [e.g., Bram's thesis; MC's CFH data]. What we do know about the early language environment is that, while child-directed speech is relatively infrequent throughout the day [@casillas2020early], shared caregiving practices and a near-universal fondness for social interaction with young children ensure that speech directed to children comes from all types of speakers: women, men, and other children [@bunceURcrosscultural; @casillas2020early]. While speech from women predominates in young children's speech environments on Rossel, as it does elsewhere [@bergelsoncasillas2019what; @bunceURcrosscultural], there is a significant and marked increase in child-directed speech from other children as infants get older [@bunceURcrosscultural; @casillas2020early]. This increase in child-directed speech from other children is attributed to the fact that, starting around age 2, children often spend large swaths of the day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 people) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].

@@

The basic structure of NWR involves the participant hearing a non-word, like "bilik", which they need to repeat back without changing anything that is contrastive in the language. For instance, in English, a response of "bilig" or "pilik" will be scored as incorrect; a response "biilik", where the vowel is lengthened without change of quality would be scored as correct, because English does not have a pure length contrast.

@@

- Does the frequency of sounds across languages predict NWR? Are rarer sounds more often substituted by commoner sounds?
- How does score change as a function of item length in number of syllables?
- Is individual variation attributable to child age, sex, birth order, monolingual status, and/or parental education?

@@

@@

We visited four different hamlets once, and attempted to test all eligible children present at the time, to prevent the items 'spreading' through hearsay. In the first hamlet, we tested children in five different places, with some children being tested inside a house and others tested on the veranda. The complete list of places and the ways in which they met the desiderata mentioned above can be found in the raw data, available from online supplementary materials.

@@

The first phase was making sure the child understood the task. This was explained orally and the first training item was presented. Often, children froze and did not say anything. If this happened, then we followed this procedure. First, the informant insisted. If the child still did not say anything, the informant asked the child to repeat a real word, and another, and another. If the child could repeat these correctly, then we provided the recorded training item over headphones again. Most children successfully started repeating the items presented over headphones at this point; a few further needed the local informant to model the behavior (i.e., they would hear the item again, and she would say it; then we would play it again, and ask the child to say it). A small minority still failed to repeat the item after hearing it over headphones. If that occurred, we tried with the second training item, at which point some children got it and could continue. A small minority, however, failed to repeat this one, as well as the third training item, in which case we stopped the test altogether (see Participants section for exclusions).

@@

read.csv("nwr-demo.csv",header=T,sep=",")->demo
#colnames(demo)

#lines no longer needed
#if family is missing, attribute each child to a different one
#demo$familyID=as.character(demo$familyID)
#demo$familyID[is.na(demo$familyID)]<-paste("k",demo$ID[is.na(demo$familyID)])
#recalc age
#demo$age.rounded=as.numeric(difftime(as.Date(as.character(demo$Date),format="%Y%m%d"),as.Date(as.character(demo$DOB.ISO),format="%Y%m%d"),unit="weeks"))/52.25
write.table(demo,"nwr-demo.csv")

@@

A total `r dim(demo)[1]` children were tested, from `r length(levels(factor(demo$familyID)))` of families, in four hamlets. Some children could not be included for the following reasons: refused participation or failed to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn\xeat repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`). In addition, `r sum(demo$Age >12 | demo$included=="too old",na.rm=T)` teenagers were tested to put younger children at ease; their data is not included in analyses below. The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged `r round(mean(inc$age.rounded),2)` years (range `r round(min(inc$age.rounded),2)`-`r round(max(inc$age.rounded),2)` years). Among these, there were `r sum(inc$monolingual=="yes", na.rm=T)` children exposed only to Yélî in the home, `r sum(inc$monolingual=="no", na.rm=T)` children who were also exposed to another language in the home, and `r sum(is.na(inc$monolingual))` for whom this information is missing. Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years; `r sum(is.na(inc$matEd))` children had this information missing).[^1] In terms of birth order, `r birthorder.tab[1]` were first borns, `r birthorder.tab[2]` second,  `r birthorder.tab[3]` third,  `r birthorder.tab[4]` forth,  `r birthorder.tab[5]` fifth, `r birthorder.tab[6]` sixth, and  `r birthorder.tab[7]` did not have this information.

@@

```{r overall}
aggregate(trnsc$correct*100,by=list(trnsc$id),mean)->overall
aggregate(trnsc$phon_score*100,by=list(trnsc$id),mean)->phon_score
aggregate(trnsc$nld*100,by=list(trnsc$id),mean)->nld
overall=cbind(overall,phon_score$x,nld$x)
overall=merge(overall, demo[,c("ID","Sex","Age","birthOrder","matEd")], by.x="Group.1",by.y="ID", all.x=T)
colnames(overall)<-c("ID","Word","Phoneme","NLD","Sex","Age","BrthO","matEd")

@@

### Patterns in NWR mispronunciations

@@

@@

```{r bylength}

aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$type),mean)->by_length
aggregate(trnsc$phon_score*100,by=list(trnsc$id,trnsc$type),mean)->phon_score
aggregate(trnsc$nld*100,by=list(trnsc$id,trnsc$type),mean)->nld
by_length=cbind(by_length,phon_score$x,nld$x)
by_length=merge(by_length, demo[,c("ID","Sex","Age","birthOrder","matEd")], by.x="Group.1",by.y="ID", all.x=T)
colnames(by_length)<-c("ID","Length","Word","Phoneme","NLD","Sex","Age","BrthO","matEd")

aggregate(trnsc$correct*100,by=list(trnsc$type),mean)->by_length_no_child
colnames(by_length_no_child)<-c("Length","Word")
by_length_no_child=by_length_no_child[-1,]
length_r=cor.test(by_length_no_child$Length,by_length_no_child$Word)

xl_fr=glmer(correct~avg_fr_s+age.rounded_s+(1|file) + (1|target),data=trnsc,family="binomial")
mycoeff=round(summary(xl_fr)$coefficients,3)

@@

@@

<!--M2A: I think a nice way to end this bit would be very briefly (~1-sentence) linking this input scenario to the relevance of looking at NWR (à la Cristia 2020 DevRev). What do you think?-->
<!-- a2m maybe we want to bring that up in the discussion? I haven't found a good place to introduce that paper, in part because right now we have these three purposes that seem sort of unconnected with the goals of that paper... -->
